---
title: "PA1_template"
author: "Ranjan Kumar"
date: "13 November 2015"
output: html_document
---

# R Markdown document for Peer Assessment 1.

## Summary
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

### Data Processing
1. Download data and Unzip it

```{r,echo = TRUE}
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  #download.file(fileUrl, destfile = "activity_monitoring.zip", method = "curl")
  #unzip("activity_monitoring.zip")
```

2. Load the data into R for Analysis
``` {r,echo = TRUE}
  activity <- read.csv("activity.csv",header = TRUE)
  activity$date <- as.Date(as.character(activity$date),"%Y-%m-%d")
```

3. Convert *data frame* into *data frame tbl* and Convert **Date** variable as Date datatype
``` {r,echo = TRUE}
  library(dplyr)
  activity <- tbl_df(activity)
```

4. Exploration of dataset.
``` {r,echo = TRUE}
  summary(activity)
```

5. Filter *missing value* **NA** from dataset and then Calculate the total number of steps taken per day
``` {r, echo = TRUE}
  steps_by_day <- activity %>% 
                        filter(!is.na(steps)) %>%
                        group_by(date) %>% 
                        summarize(tot_steps = sum(steps))
  steps_by_day
```

6. Histogram of the total number of steps taken each day
```{r}
  hist(steps_by_day$tot_steps,col="red",xlab = "Steps", main = "Total number of steps taken each day")
```

7. Mean & Median of *Total number of steps taken per day*
``` {r}
  steps_by_day %>% summarize(mean = mean(tot_steps), median = median(tot_steps))
```

8. Time series plot of Average Steps Taken
``` {r, echo = TRUE}
  avg_steps_by_interval <- activity %>% 
                              filter(!is.na(steps)) %>%
                              group_by(interval) %>% 
                              summarize(avg_steps = mean(steps))
  
  plot(avg_steps_by_interval$interval,avg_steps_by_interval$avg_steps,type = "l",xlab = "Interval", ylab = "average number of steps taken",main = "Time Series Plot of Avg Step")
```

9. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
``` {r, echo=TRUE}
  avg_steps_by_interval %>% arrange(desc(avg_steps))
  
```

10. Calculate and report the total number of missing values in the dataset 
``` {r,echo=TRUE}
    activity %>%
            filter(is.na(steps)) %>%
            summarise(rows_with_missing_value=n())
  
```

11. Seperate complate & incomplete observations into two groups
```{r, echo = TRUE}
  activity_complete <-  activity %>% filter(!is.na(steps)) 
  activity_NA <- activity %>% filter(is.na(steps))
```

12. Filled **NA observations**  with **Average steps taken in Interval**
i.e. Merge incomplete observations with *Average Steps By Interval*
``` {r,echo = TRUE}
  activity_NA_filled <- merge(activity_NA,avg_steps_by_interval,by = "interval") %>%
                        select(avg_steps,date,interval)
```

13. Align variable name of *activity_NA_filled* with data frame *activity_complete* followed by rbind
```{r,echo=TRUE}
  names(activity_NA_filled) <- c("steps","date","interval")
  activity_NEW <- rbind(activity_complete,activity_NA_filled)
  summary(activity_NEW)
```

14. Cleanup temporary data frames
``` {r,echo=TRUE}
  rm("activity_NA_filled","activity_NA","activity_complete")
```

15. Calculate the total number of steps taken per day from new dataset
``` {r, echo = TRUE}
  steps_by_day_NEW <- activity_NEW %>% 
                        group_by(date) %>% 
                        summarize(tot_steps = sum(steps))
  steps_by_day_NEW
```

16. Mean & Median of *Total number of steps taken per day*
``` {r}
  steps_by_day_NEW %>% summarize(mean = mean(tot_steps), median = median(tot_steps))
```

17. Create a new factor variable in the dataset with two levels – weekday and weekend
```{r,echo=TRUE}
  weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
  activity_NEW <-mutate(activity_NEW,days = factor((weekdays(activity_NEW$date) %in% weekdays1), 
                                                   levels=c(FALSE, TRUE), 
                                                   labels=c('weekend', 'weekday')))
  
  activity_NEW
```

18. Time series plot of Average Steps Taken
``` {r, echo = TRUE}
  library(ggplot2)
  avg_steps_by_interval_NEW <- activity_NEW %>% 
                              group_by(days,interval) %>% 
                              summarize(avg_steps = mean(steps))
  
  g <- ggplot(avg_steps_by_interval_NEW,aes(interval,avg_steps))
  g + geom_line() + labs(x="Interval") + facet_grid(. ~ days) + facet_wrap( ~ days, ncol=1) + labs(y="Average Steps") + labs(title= expression("Time Series Plot"))

```
